Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-16 paper/paper.tex
A better alternative fn for tectal ligands
Graphing updates
More WIP

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that \textbf{i)} All
receptor-ligand signalled interactions are repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecligexp}.

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json  configs/simpler/e_wt_fig1.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300
steps. Fig.\,\ref{f:chemo}A indicates the colour code used for retinal cell
position; dorsal retinal cells are indicated by green, nasal cells by
red. Lightness gives position along the relevant axis so that cells at the
dorsal-nasal corner are yellow and those at the ventral-transverse corner are
black. Fig.\,\ref{f:chemo}B shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina, as shown in Fig.\,\ref{f:chemo}A. Nasal cell axons should be
arranged caudally; dorsal axons laterally; axons should be evenly spaced and
the pattern should span the entire tectum. Figs.\,\ref{f:chemo}C
and \ref{f:chemo}D show, respectively, the positions of branches (i.e.~growth
cones) and axon centroids at $t=300$. Topographic ordering is
correct. Opposing gradients have sorted the axons along the two axes to form a
grid of cells. However, the arrangement does not span the full
tectum. Fig.\,\ref{f:chemo}E shows selected axons (one from the centre and one
from each corner of the retina) and their position history.
Fig.\,\ref{f:chemo}F shows two pattern metrics; the sum of squared distance
errors of the axon centroids and the number of crossings in the 'fish net'
plot of Fig.\,\ref{f:chemo}D which gives an indication of the disorder in the
arrangement. The expected number of crossings is 0, as in
Fig.\,\ref{f:chemo}B.

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:chemo}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.2 x)
\end{equation}

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig2.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_GJ_wt_fig.png}
\caption{Gradient following model with axon-axon interactions.}
\label{f:wt}
\end{figure}

@@

Introductory sentence.

\textbf{FIXME: Find the expression functions for the tectum which give a
decent WT result. Perhaps I already did this? Experiment with these and
axon-axon with smaller radius of interaction.}

We now describe the behaviour of a model which has only the gradient-based
chemoaffinity effect and no axon-axon interaction or border effect
(i.e.~$m_2=m_3=0$). This model was run with $n=8$ branches per RGC axon. All
receptor-ligand signals induced repulsive interactions (i.e.~gradient
descent). Receptor and ligand expressions were as shown in Fig.\,\ref{f:ex},
noting that tectal ligand expression was of exponential form
(Eq.\,\ref{e:tecligexp}). Fig.\,\ref{f:ch}A gives a colour map of retinal
ganglion cell soma locations, with the colour red encoding $x$ position on the
temporal-nasal axis (with red signifying nasal positions). Green encodes $y$
position on the dorso-ventral axis.
%
Fig.\,\ref{f:ch}B shows the expected final location of each RGC axon,
according to experimental observations, in a `fishnet' plot which draws lines
between adjacent axons giving a visual indication of the topology of the
arrangement.
%
Fig.\,\ref{f:ch}C shows the position of
400 branches at $t=500$ (in arbitrary time units). Considering each
coordinate axis separately, each branch interacts differentially with two
opposing gradients. Branches which settle nearest the rostral edge of the
tectum interact more strongly with the gradient $L_0$ than with $L_2$
(Fig\,\ref{f:ex}C). A similar pair of gradients separate branches along the
M-L tectal axis. The branches become positioned on a grid due to the
discretization of the tectal ligand gradient (as the number of branches and
tectal elements tends towards infinity, the arangement tends towards a
continuous map---see Fig.\,S1 \textbf{FIXME}).

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_chemoonly_wt_chemoonly.png}
\caption{Chemoaffinity model}
\label{f:ch}
\end{figure}

Fig.\,\ref{f:ch}D shows a fishnet plot of the RGC axon centroids; the mean
location of $n$ branches per axon. Because all $n$ branches for each RGC have
identical receptor expression and they experience only the chemoaffinity
effect, each branch attains almost the same final position and so the centroid
graph strongly resembles Fig.\,\ref{f:ch}B. Fig.\,\ref{f:ch}D indicates that
the topology of the arrangement resembles that of the experimental result, but
the area over which the branches arrange themselves is limited to about 50\%
of the tectum. It is possible to tailor the tectal ligand expression, whcih
adjusts the relative contributions of the opposing interactions, to tune the
final map. This allows the chemoaffinity model to attain a good match to the
experimental arrangement (Fig.\~S2 \textbf{fixme}). However, the match is
highly dependent on the form of receptor/ligand expression in both retinal and
tectal cells.

\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. Row A: wildtype. Row B: 90 degree
rotation. Row C: 180 degree rotation. Row D: Graft swap}
\label{f:chemosurg}
\end{figure}

>>>>

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that: \textbf{i)} all
receptor-ligand signalled interactions are repulsive ($F_i=-1, \forall i$), as
for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecligexp}.

@@

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300
steps.

Fig.\,\ref{f:ch}A indicates the colour code used for retinal cell position;
dorsal retinal cells are indicated by green, nasal cells by red. Lightness
gives position along the relevant axis so that cells at the dorsal-nasal
corner are yellow and those at the ventral-transverse corner are
black. Fig.\,\ref{f:ch}E shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina, as shown in Fig.\,\ref{f:ch}A. Nasal cell axons should be arranged
caudally; dorsal axons laterally; axons should be evenly spaced and the
pattern should span the entire tectum.  Fig.\,\ref{f:ch}B--D show `fish net'
plots of the centroid of the branches of each axon at three timepoints. In
these plots, lines are drawn between axons that are expected to arrange
themselves adjacently to one another. The number of line crossings in this
plot gives an indication of the disorder in the arrangement. Fig.\,\ref{f:ch}G
shows the positions of the individual branches (i.e.~growth cones) at $t=300$
that contribute to the the plot in Fig.\,\ref{f:ch}D. By this time, the
topological order matches that shown in Fig.\,\ref{f:ch}E. Opposing gradients
have sorted the axons along the two axes to form a grid of cells. However, the
arrangement does not span the full tectum. Fig.\,\ref{f:ch}F shows two pattern
metrics; the sum of squared distance errors of the axon centroids and the
number of crossings in the fish net plot of Fig.\,\ref{f:ch}D. The expected
number of crossings is 0, as in Fig.\,\ref{f:ch}E. Fig.\,\ref{f:ch}H shows
selected axons (one from the centre and one from each corner of the retina)
and their position history.

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model.}
\label{f:ch}
\end{figure}

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:ch}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.1 x)
\end{equation}

The result at $t=300$ is shown in Fig.\,\ref{f:chalt}A and B. The axon
centroids are arranged across the full tectal surface.  The arrangement is
slightly disordered around the middle of the tectum. This is caused by
numerical artefacts from the discretization of the ligand expression and the
fact that in the centre, the difference between the opposing receptor-ligand
signals is very small.  Fig .\,\ref{f:chalt}C shows the error metric, which is
closer to 0 with the new ligand expression.

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3.png}
\caption{The pure gradient-following model (alt exponential).}
\label{f:chalt}
\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./images/j4_eE_GJ_wt_fig.png}
%\caption{Gradient following model with axon-axon interactions.}
%\label{f:wt}
%\end{figure}

@@

This section now at start of results

\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. Row A: wildtype. Row B: 90 degree
rotation. Row C: 180 degree rotation. Row D: Graft swap}
\label{f:chsurg}
\end{figure}
